The breath shape controls intonation of mouse vocalizations

Version 1

Abstract

Abstract

Intonation in speech is the control of vocal pitch to layer expressive meaning to communication, like increasing pitch to indicate a question. Also, stereotyped patterns of pitch are used to create distinct “words”, like the ten sounds in the murine lexicon. A basic tone is created by exhalation through a constricted laryngeal voice box, and it is thought that more complex utterances are produced solely by dynamic changes in laryngeal tension. But perhaps, the shifting pitch also results from altering the power of exhalation. Consistent with the latter model, we describe that intonation in many adult murine vocalizations follows deviations in exhalation and that the brainstem vocalization central pattern generator, the iRO, can create this breath pattern. Consequently, ectopic activation of the iRO not only induces phonation, but also the pitch patterns that compose most of the vocalizations in the murine lexicon. These results reveal a novel brainstem mechanism for intonation.

Introduction

Modulation of the frequency of produced sound, perceived as pitch, creates meaning within words or phrases through intonation (Prieto et al., 2015). For example, in English, an increasing pitch is used to indicate a question or stress importance and a decreasing pitch communicates a declaration. Additionally, the concatenation of specialized sounds with variation in pitch, like syllables, composes the diverse repertoire of words (Poeppel et al. 2020). Two key pieces of the phonation system are the larynx (the “voice-box”) and the breathing muscles (Berke et al. 2009, Laplange et al., 2018). Succinctly, the breathing muscles drive airflow through a narrowed larynx to produce a basic vocalization (Finck et al., 2009). The speed of the airflow through the larynx dictates the fundamental frequency of the tone, so changes in either the forcefulness of the breath exhalation or the extent of laryngeal closure can both, presumably, alter the pitch (Kelm-Nelson et al., 2018, Herbst C.T., 2016, Mahrt et al., 2016). While control of the size of the laryngeal opening is well established as a mechanism to regulate the dynamic changes in pitch for human to rats and mouse vocalizations (Poeppel et al. 2020, Johnson et al., 2010, Riede et al., 2017), the contribution of exhalation itself remains to be carefully defined. In fact, it is presumed that the forcefulness of expiration only modulates the vocal amplitude or loudness (Riede T., 2011, Riede T., 2013) This perception stems from the airflow of the rodent breath not strongly predicting the pitch. Yet paradoxically, an injection of air below the larynx to enhance airflow increases pitch (Riede T., 2011). This incongruity even extends into songbirds, a leading vocalization model system (Suthers et al., 2002, Schmidt et al., 2014, Plummer et al., 2008, Goller et al., 2004). Here, we seek to resolve this inconsistency by taking advantage of the experimental, behavioral, and genetic approaches in the mouse (Yackle K., 2023). If two independent variables are used to alter pitch, like the larynx and the breath airflow, then the interplay would enhance the ability to produce a diverse repertoire of sounds and thereby enable a broader lexicon.

The medullary brainstem possesses at least two means that might account for the two control points proposed above, laryngeal diameter and exhalation strength. First, direct modulation of laryngeal premotor and motor neurons in the retroambiguus (RAm) modulates the size of the laryngeal opening (Kelm-Nelson et al., 2018, Hage S.R., 2009). And second, the vocalization central pattern generator we recently described, called the intermediate Reticular Oscillator (iRO), induces coordinated changes in the expiratory airflow and laryngeal closure (Wei et al., 2022). For example, during neonatal cries, the iRO oscillates exhalation strength and larynx activity to time the syllable sounds. Thus, the RAm provides a mechanism to modulate pitch by controlling laryngeal diameter independently from the iRO altering the tone by dictating the extent of the breath expiratory airflow. While the contribution of RAm in adult phonation has been established (Jürgens U., 2002), the role of the iRO remains undefined.

Here, we describe the coordinated changes in breath airflow and pitch in the ten vocalizations of the adult murine lexicon (Grimsley et al., 2011). We describe that the modulation of pitch for the different vocalizations either correlates or anticorrelates with the changes in exhalation. These results support a model in which two independent mechanisms involving changes in laryngeal opening or airflow control tone. Using anatomical, molecular, and functional approaches, we demonstrate that the iRO vocal central pattern generator drives changes in breath expiratory airflow to pattern pitch and produce seven of the ten vocalizations in the endogenous lexicon. These data resolve the prior paradoxical role of exhalation and show it can directly control pitch. Additionally, we establish the iRO as a mechanistic basis for intonation. And lasty, these results generalize the crucial role of the iRO in phonation across developmental stages, and we presume across species.

Results

Vocalizations are produced by a program coupled to breathing

It is possible that the ten murine ultrasonic vocalizations (USVs) defined by unique pitch patterns (Grimsley et al., 2011) are formed by distinct breaths or as substructures nested in a common breath. Prior work has suggested the latter (Sirotin et al., 2014). To expand upon this, we simultaneously measured breathing and USVs by customizing the lid of a whole-body plethysmography chamber to accommodate a microphone. Male mice in the chamber were exposed to fresh female urine and robustly sniffed and vocalized for the first 5-10 minutes of the recording at a peak rate of about 4 events per second (n = 6) (Fig. 1A-B). A vocalization was classified as a narrow-band sound in the 40-120 kHz ultrasonic frequency range during a single breath (Fig. 1A). The rate of vocalization breaths was typically between 5-10 Hz (Fig. 1C) and mostly occurred during episodes of rapid sniffs (∼10 Hz) (Fig. 1A and B), as previously reported (Sirotin et al., 2014, Castellucci et al., 2018) When compared to neighboring breaths, vocalization breaths had slightly larger inspiratory and perhaps expiratory airflow despite similar durations of each phase (Fig. 1D and E). These data reveal that a vocalization breath appears mostly like a normal breath, but with the addition of a nested sound pattern. This led us to hypothesize that a distinct sub-program is activated within a breath to generate a vocalization.

The full repertoire of vocalizations occurs within a normal appearing breath.

A, Male mice exposed to female urine produce ultrasonic vocalizations (USV) at about 75 kilohertz (top) that coincide with the expiratory airflow (E, arbitrary units) of the breath cycle (bottom). Red box indicates the length of the USV. A bout of vocalizations contains breaths with USVs (red) interspersed among sniff breaths (black). B, Rates of breathing (black) and USV production (red). Exposure to female urine at time 0, n=6 mice. C, Histogram of the instantaneous frequency of breaths with and without USVs from n=6 animals. p-value 0.98; two-way ANOVA. D, Scatter plot of the inspiratory (Ti) and expiratory time (Te) for USV (red) and non-USV (black) breaths from n=1 representative animal. Right, bar graph of mean ± SEM of Ti, Te, and the ratio for n=6. Each dot is the mean from each animal. p-values 0.40, 0.18, and 0.25; paired t-test. E, The breath peak inspiratory (pif) and expiratory (pef) airflow represented as in D. p-values 0.01, 0.27; paired t-test. F, Bar graph (mean ±SEM) of the percent of total USVs for each type from n=6 mice. Each dot is the mean from each animal.

The adult murine lexicon is composed of at least 10 USV types that are defined by different, but stereotyped patterns of pitch. One breath can contain multiple syllables, which we define as a continuous USV event within a breath (Fig. 1F and S2-3). A pre-trained convolutional neural network (CNN) was used to classify USVs into different types based on changes in pitch (Fonseca et al., 2021) and the on- and offset of each vocalization was overlayed upon the corresponding breath airflow (Fig. S1-3). Vocalizations began and ended throughout expiration (Fig. S1), and the most common tended to start near the onset of exhalation and ended shortly there-after (like the up frequency modulated, step down, flat, and short types) (Fig. S2). Vocalizations with more intricate changes in pitch had more variable times of on- and offset (like complex, chevron, two step, multi, step up, down frequency modulated) (Fig. S3). And lastly, when the vocalizations occurred late in expiration, the duration of this breath phase was prolonged (Fig. S4). The bias of USV timing by the breath, combined with the USV modulation of breath length demonstrate these programs are independent but reciprocally coupled.

Two mechanisms create the changes in pitch pattern

Fluctuations in airflow through the larynx produce changes in the sound’s pitch. For example, augmenting airflow through the explanted rodent larynx increases pitch (Mahrt et al., 2016). We proposed two potential models that would explain how the laryngeal airflow is modulated to form the distinct USV types in the murine lexicon: one based on the strength of exhalation pushing air through the larynx, and another based on the diameter of the laryngeal opening (Fig. 2A). According to the first model which we term positive intonation, if the pitch changes mirror the modulation of the breath expiratory airflow, the plethysmography airflow and pitch will simultaneously fluctuate (Fig. 2A, left). In the opponent model, negative intonation, a narrowed larynx used to increase pitch would impede the overall expiratory airflow such that pitch is anticorrelated with plethysmography airflow (Fig. 2A, right). Note, these models can form similar airflow patterns, but predict opposite relationships to pitch.

The ten types of USVs are produced by at least two mechanisms that modulate airflow.

A, Two models to control the USV pitch by changing the speed of airflow through the larynx. Left, modulation in expiratory airflow drives the change in pitch (positive intonation, blue). Right, the change in pitch anti-correlates with airflow suggests closed larynx produces sound (negative intonation, green). B, Left, example of the expiratory airflow and pitch for a down frequency modulated (fm) USV. Middle, magnification of airflow and sound. The scale of airflow is not displayed. The time of breath airflow from expiration onset during the USV is color coded blue to white. Note, the change in pitch mirrors airflow, consistent with the ‘breath modulation’ mechanism (annotated as “+”). Scatter plot of instantaneous expiratory airflow and pitch for a single USV and the correlation (line, r). Box and whisker plot of n=40 down fm correlation coefficients (r). C, Representative expiratory airflow and pitch and box and whisker plot of all r values for complex (n=165) vocalizations. D, Up frequency modulated (fm) vocalization represented as in B (n=589). Note, the change in pitch opposes the airflow, consistent with the ‘laryngeal closure’ mechanism. (annotated as “-”). E, Two step vocalization represented as in B (n=61). The airflow for each unique USV element is uniquely color coded as green, blue, or purple. Note, the change in pitch for two components correlates and one anticorrelates. This is consistent with both mechanisms being sequentially used. Annotated as mixed blue and green box and whisker plot. F, Box and whisker plot of correlation coefficients (r) for step up (n=40), multi (n=58), step down (n=293), and chevron (n=99). G, Model schematic of the mechanisms used to control pitch for the various USV types.

We assessed each model by calculating the correlation coefficient (r) between instantaneous expiratory airflow and the corresponding USV fundamental frequency. Down or up frequency modulated USVs were positively or negatively correlated, respectively (median r = 0.62 and -0.46, Fig. 2B and D). These simple USVs reflected the two proposed mechanisms to alter pitch, positive and negative intonation (Fig. 2A). Six of the other ten USVs types had positively shifted intonation, the chevron, complex, step down, multi, and two step types (median r = 0.32, 0.31, 0.28, 0.24, and 0.19 respectively), and many of the step up were negatively biased (median r = -0.03) (Fig. 2C, E, F). For many of these USV types, it appeared that a portion of the USV pattern correlated with the expiratory airflow, while the other part(s) were un- or anticorrelated (e.g., the two step, Fig. 2E). This suggests that the pitch is patterned by switching between positive and negative intonation mechanisms within the breath (Fig. 2G). The remaining two USV types (flat and short) had different breath shapes, which resulted in a wide range of r values (Fig. 4). In summary, all these results support the hypothesis that a vocalization pattern generator must integrate with and even control the breath airflow as a key mechanism to produce various USV types in the murine lexicon (Fig. 2G).

The iRO resides within the adult brainstem phonation circuit

Murine vocalizations are innate and stereotyped (Fig. S1 and 2) which predicts they are generated by a vocal central pattern generator (CPG). The similarities between the positively correlated USV types and the neonatal cry vocalizations produced by a vocal CPG known as the intermediate Reticular Oscillator or iRO (Wei et al., 2022) suggests the iRO is involved in generating adult USVs. However, the iRO has yet to be identified in adult mice.

The iRO is molecularly defined in the neonate by the co-expression of Preproenkephalin (Penk) and Vesicular glutamate transporter 2 (Vglut2) and is anatomically localized to the medullary ventral intermediate Reticular Formation (iRT) directly medial to the compact nucleus ambiguus17,19 (Wei et al., 2022). This general region has been dubbed the Post inspiratory Complex (PiCo) given its involvement in post-inspiration, including behaviors like swallowing (Anderson et al., 2014, Huff et al., 2023). We determined that the iRO molecular and anatomical features exists in adults in two ways. First, we generated triple transgenic mice that label Penk+Vglut2+ neurons and the derived lineages with tdTomato (Penk-Cre; Vglut2-Flp; Ai65) (Fig. 3A). And second, we stereotaxically injected the iRO region of Penk-Cre; Vglut2-Flp mice with a Cre and Flp dependent reporter adeno-associated virus (AAV CreONFlpON-ChR2::YFP) (Fig. 3B-C). Consistent with the definition of the iRO in neonatal mice, tdTomato+ and YFP+ Penk+Vglut2+ neurons were found in the iRT adjacent to the compact nucleus ambiguus (Fig. 3A-C). These results demonstrate that the ventrolateral medulla of adult mice contains neurons with the molecular and anatomical identity of the iRO.

Anatomically and molecularly defined iRO neurons form a brainstem phonatory circuit.

A, Labeling of Penk+Vglut2+ neurons in the iRO anatomical region in adult Penk-Cre;Vglut2-Flp;Ai65 mice (CreONFlpON-tdTomato) (observed in n=5 mice). The iRO region is defined as medial to the compact nucleus ambiguus (cNA, ChAT +) in the ventral intermediate Reticular Tract (iRT). Note, the cNA is filled with tdTomato labeled axons. Cell bodies marked with arrowhead. B, Bilateral stereotaxic injection of AAV CreONFlpON-ChR2::EYFP into the iRO anatomical region of Penk-Cre;Vglut2-Flp adult mice. C, Magnified boxed region in B. Arrowheads label neuron soma quantified right (n=3). D, Axons of EYFP expressing iRO neurons from B in the retroambiguus (RAm) where laryngeal motor neurons are located. E, Axons of EYFP expressing iRO neurons from B in the breathing pacemaker. F, Unilateral retrograde AAV CreON-EYP (AAVrg) stereotaxic injection into the iRO region in Vglut2-Cre adults (n=3). Glutamatergic neurons were identified in the contralateral (contra.) and ipsilateral (ipsi.) midbrain periaqueductal gray (PAG). Anatomical regions of the PAG: dorsomedial (dm), dorsolateral (dl), lateral (l), ventrolateral (vl) nearby to the dorsal raphe nucleus (DRN) and surrounding the cerebral aqueduct (Aq). G, Quantification of glutamatergic PAG neurons in each region demarcated in F, ns = not statistically significant; two-way ANOVA with Sidka’s post-hoc test. H, Model schematic of the iRO as a central component of the brainstem phonation circuit to convert a vocalization “go” cue from the PAG into a motor pattern.

Neonatal iRO neurons are presynaptic to the kernel of the breathing, the pacemaker for inspiration (preBötzinger Complex, preBötC) (Smith et al., 1991) and premotor to multiple laryngeal and tongue muscles. We traced the YFP+ axons of Penk+Vglut2+ neurons (Penk-Cre; Vglut2-Flp and AAV CreONFlpON-ChR2::YFP) and found they elaborated within the nucleus ambiguus (NA) and retroambiguus (RAm) where laryngeal premotor and motor neurons localize (Fig. 3A, D), the breathing pacemaker (Fig. 3E), and the hypoglossal (tongue) motor nucleus (Fig. 3E). The projection patterns of these Penk+Vglut2+ neurons provide additional evidence that these adult neurons maintain the same connectivity properties as the neonatal iRO neurons, indicating they can control the key elements for vocalization: the breath airflow and larynx.

In adult mice, vocalizations have been triggered by activation of the midbrain periaqueductal gray (PAG), namely glutamatergic neurons in the ventrolateral subregion (Michael et al., 2020, Chen et al., 2021, Tschida et al., 2019). To assess if the iRO region is positioned downstream of the ventrolateral PAG, we unilaterally injected Vglut2-Cre mice with a CreON-ChR2::YFP expressing retrograde traveling AAV (AAVrg) (Fig. 3F). Among the labeled brain regions, we found YFP+ neurons selectively in the phonation region of the midbrain PAG. To our surprise, neurons from the ipsi- and contralateral PAG projected to the iRO region in nearly equal numbers (Fig. 3G). These molecular, anatomical, and neural morphology characterizations reveal that the iRO exists in adults and is embedded within the brainstem phonation network (PAG → iRO → the preBötC, NA, RAm, hypoglossal) (Fig. 3H).

Ectopic activation of the putative iRO induced vocalization

If these labeled Penk+Vglut2+ neurons are indeed the iRO, we anticipated that ectopic activation would induce vocalization. We tested this in two ways. First, we generated Penk-Cre;Vglut2-Flp;CreONFlpON-ReaChR triple transgenic mice which express the red-shifted Channel Rhodopsin in Penk+;Vglut2+ neurons and the derived lineage (ReaChR mice) and second, we stereotaxically injected the AAV CreONFlpON-Channel Rhodopsin2::YFP (ChR2) into the iRO region of Penk-Cre;Vglut2-Flp mice. In both instances we implanted optic fibers above the iRO bilaterally to further localize neural activation (Fig. 4A and S5A). In both experimental regimes, ectopic light activation of the Penk+Vglut2+ neurons induced bouts of vocalizations where the breathing rate was entrained by the frequency of stimulation (Fig. 4A, S5A, S5I). Most bouts and the breaths within contained vocalizations (Fig. 4B and S5B), and the amplitudes of all elicited breaths were significantly increased (Fig. S5J). Some AAV-ChR2 mice showed broad band vocalizations, while others did not vocalize, likely due to incomplete labeling (n=5/9). Additionally, the ReaChR animals without vocalizations were found to have “off target” optic fiber implants (n=2/6). Taken together, these data are consistent with the notion that the iRO is sufficient to induce phonation via control of both breath airflow and laryngeal opening, just as it does in neonatal cries.

Ectopic activation of the iRO evokes airflow correlated USV types and switches the relationship of the anti-correlated types.

A, Optogenetic activation of the iRO region in Penk-Cre;Vglut2-Flp;CreONFlpON-ReaChR mice evokes USVs (blue box, 5 Hz stimulation). USVs occur during, or shortly after laser onset. B, Percentage of stimulation bouts containing at least one USV and the percentage of breaths within the stimulation window containing a USV. C, Percentage of optogenetically evoked (blue) or endogenously occurring (gray) syllables that are up fm of down fm. **** p-value< 0.001; two-way ANOVA with Sidak’s post-hoc test, p > 0.05 for all other types. D, Box and whisker plot of the correlation coefficient between breathing airflow and pitch (r) for all opto evoked (n=395) and endogenous (n=1850) USVs from n=4 opto and n=6 endogenous mice. **** p-value < 0.001; Mann-Whitney test. E, Left, example of the expiratory airflow and pitch for an optogenetically evoked up fm USV. Middle, magnification of airflow and sound. Time of breath airflow during the USV is color coded blue to white. Scatter plot of instantaneous expiratory airflow and pitch for the single USV. Compare to endogenous up fm USV in Fig. 2D. Right, box and whisker plot of correlation coefficients (r) for each optogenetically evoked and endogenous up fm USV (n=15 vs. 589). F, Step up USV as in E. box and whisker plot of correlation coefficients (r) for each optogenetically evoked and endogenous step up USV (n=15 vs. 40). G, r value box and whisker plots for the remaining optogenetically evoked USV types (down fm n=242 vs. 40, step down n=38 vs. 293, flat n=34 vs. 337, short n=27 vs. 168, and chevron n=10 vs. 99 from n=4 opto and 6 endogenous mice. E-G, two way ANOVA with Sidak’s post-hoc test for two way comparisons was used; all p-values >0.05. H, Schematic illustrating the two mechanisms to pattern the USV pitch. Left, the reciprocal connection between the iRO and breathing pacemaker patterns the USVs with a positive correlation between pitch and airflow. Right, the retroambiguus (RAm) control of the larynx dictates the anti-correlated USV types. Middle, the combination of these two mechanisms within a single breath create additional USV patterns.

To demonstrate the specialization of the iRO neurons for vocalization and the inability of modulated breathing alone to elicit USVs, we performed several additional control experiments. First, to ensure that just stimulation of breathing is insufficient to elicit vocalization, we optogenetically excited the glutamatergic preBötC neurons (Vglut2-Cre with AAV CreON-ChR2). Indeed, we found that, although breathing sped up, optogenetic stimulation never elicited vocalizations (Fig. S5C, H, I). And second, to determine if the ability to elicit vocalizations was generalizable to other neural types in the iRO anatomical region, we activated Penk+, µ-opioid receptor+Vglut2+, Tachykinin 1+, and Vesicular GABA transporter+ neurons and found that vocalizations were never induced upon light stimulation, although breathing was altered in various ways (Fig. S5). In summary, these data functionally demonstrate the existence of Penk+Vglut2+ iRO neurons in adult mice and their ability to create vocalizations by modulating both breathing and presumably the larynx.

Excitation of the iRO evoked nearly the entire murine lexicon

Above, we described that one mechanism for generating the different patterns of vocalizations was via the modulation of the breath airflow (positive intonation). Once again, this was defined as a positive correlation between expiratory airflow and pitch (Fig. 2). We hypothesized that this property stems from the iRO’s capacity to control breathing, and so we made the following predictions: 1) that the USVs evoked after stimulation would be biased to those with an endogenous positive correlation between airflow and pitch (like the down fm and step down), and 2) that any of the elicited USV types would be transformed to become more positively correlated.

We classified the evoked iRO vocalizations (Penk-Cre;Vglut2-Flp;CreONFlpON-ReaChR) with the CNN, and to our surprise, seven of the ten types of endogenous USVs were induced upon activation of the iRO (Fig. 4). The most abundant elicited USV was the down fm which, in the endogenous dataset, had the strongest positive intonation (Fig 4C and Fig. 2). Conversely, the USV with the strongest negative intonation was rarely found, up fm. These results are striking since the down fm is the least common endogenous USV and up fm is the most common (Fig. 2). These results are consistent with the first prediction where the optically evoked USV types were biased towards those with endogenous positive intonation. Beyond this, all the ectopic USVs were transformed to positively associate airflow and pitch, even when the counterpart endogenous USV was un- or anticorrelated (e.g., up fm and step up USVs) (Fig. 4E-F). This aligns with the second prediction. These data demonstrate that the iRO is sufficient to pattern nearly all USV types, and that the pitch of the induced vocalizations tightly follows the breathing airflow.

Discussion

Here we propose that the intonation that establishes the diversity of the adult murine lexicon is explained by two mechanisms, the modulation of the breath waveform and presumably the size of the laryngeal opening. First, we describe that unique vocalization types have characteristic fluctuations in the expiratory airflow, whereby some changes in pitch are strongly correlated with airflow while others are anticorrelated. These two mechanisms can even be used in the same breath to produce complex changes in pitch. To our surprise, six of the ten USV types primary used the positive intonation mechanism. These data support a novel and key role for the breathing system in the production of various types of vocalizations. Second, we show that the vocalization central pattern generator, the iRO, is sufficient to induce most of the endogenous USVs types via the modulation of the breath airflow. In contrast to the natural lexicon, the pitch of the evoked USVs is explained by positive intonation. These data imply that the iRO can produce the mechanism to pattern positive intonation, and thereby suggests that negative intonation derives from a separate neuronal component of the phonatory system. We propose these two mechanisms can be used independently or in conjunction to generate the diverse repertoire of vocalizations (Fig. 4H).

The iRO likely patterns intonation for endogenous phonation

The description of the iRO within the adult neural circuit for phonation suggests a key role in patterning the endogenous adult vocalizations. In this case, we propose that the upstream periaqueductal gray input would “turn-on” the iRO which then co-opts the breathing pacemaker and coordinates its modulation with laryngeal activity to produce and pattern the coordinated changes in breath airflow and vocal pitch (positive intonation). The iRO can do this since it is presynaptic to both the breathing pacemaker and the laryngeal motor neurons. In this case, the brief re-activation of inspiratory muscles would slow ongoing expiration, enabling bi-direction changes in airflow, and thus pitch. This type of modulation has been demonstrated in neonatal cries (Wei et al., 2022). An important next step will be to validate this supposition by correlating measurements of breathing muscle activity with pitch. Also, future studies should explore the need of the iRO in adult phonation, as anticipated from its necessary role in neonatal vocalization. None-the-less, the presence of the iRO across developmental stages implies a conserved role in innate vocalizations within the mouse and perhaps across the animal kingdom, where vocalization central pattern generators have been hypothesized and even identified in species from fish to birds to primates (Zhang et al., 2020, Chagnaud et al., 2011, Hage S.R., 2009, Kelley et al., 2020).

The iRO can autonomously produce multiple vocalization patterns

A surprising finding is that ectopic activation of the iRO produces seven of the ten vocalization types within the murine lexicon. How might this occur? One possibility is that the iRO has multiple modes which can each produce a different pattern of activity. Such a phenomena has been demonstrated in other central pattern generating systems like the crustacean stomatogastric ganglia (Marder et al., 2001, Marder E., 2012). A more likely option is that additional mechanisms of vocal modulation are layered upon a basic pattern produced by the iRO. For example, other regions with direct control of the laryngeal motor neurons within RAm would add complexity to the vocalization induced by the iRO, akin to how vocal control by the human laryngeal motor cortex is perceived (Fig. 4H) (Dichter et al., 2018, Silva et al., 2022). Here we propose that perhaps just two mechanisms (breath airflow and laryngeal opening) account for the intricacy of the murine sounds produced, and the layering of these enables a basic pitch structure within a breath to become sophisticated.

The control of breathing airflow is a novel biomechanical mechanism for intonation

Intonation is a key aspect of communication, whereby the same word or phrase could be used as a question or a statement simply by different fluctuations in pitch. Our findings describe a novel biophysical mechanism for intonation and a cellular basis. Now, the iRO or the direct modulation of breathing can serve as a starting point to map higher level components of brain-wide vocalization circuits that structure additional subliminal layers of perception in speech.

eLife Assessment

This important study examines the relationship between expiratory airflow and vocal pitch in adult mice during the production of ultrasonic vocalizations and also identifies a molecularly defined population of brainstem neurons that regulates mouse vocal production across development. The evidence supporting the study's conclusions that expiratory airflow shapes vocal pitch and that these brainstem neurons preferentially regulate expiratory airflow is incomplete and would benefit from the inclusion of additional analyses and discussion. This work will be of interest to neuroscientists working on mechanisms and brainstem circuits that regulate vocal production and vocal-respiratory coordination.

Reviewer #1 (Public Review):

Summary:

In this important work, the authors propose and test a model for the control of murine ultrasonic vocalizations (USV) in which two independent mechanisms, involving changes in laryngeal opening or airflow, control vocal tone. They present compelling experimental evidence for this dual control model by demonstrating the ability of freely behaving adult mice to generate vocalizations with various intonations by modulating both the breathing pattern and the laryngeal muscles. They also present novel evidence that these mechanisms are encoded in the brainstem vocalization central neural pattern generator, particularly in the component in the medulla called the intermediate reticular oscillator (iRO). The results presented clearly advance understanding of the developmental nature of the iRO, its ability to intrinsically generate and control many of the dynamic features of USV including those related to intonation, and its coordination with/control of expiratory airflow patterns. This work will interest neuroscientists investigating the neural generation and control of vocalization, breathing, and more generally, neuromotor control mechanisms.

Strengths:

Important features and novelty of this work include:

1. The study employs an effective combination of anatomical, molecular, and functional/ behavioral approaches to examine the hypothesis and provide novel data indicating that variations in expiratory airflow can change the pitch patterns of adult murine USV.

2. The results significantly extend the authors' previous work that identified the iRO in neonatal mice by now presenting data that functionally demonstrates the existence of the critical Penk+Vglut2+ iRO neurons in adult mice, indicating that the iRO neurons maintain their function in generating vocalization throughout development.

3. The results convincingly demonstrate that the iRO neurons encode and can generate vocalizations by modulating both breathing and the laryngeal muscles.

4. The anatomical mapping and tracing results establish an important set of input and output circuit connections to the iRO, including input from the vocalization-promoting subregions of the midbrain periaqueductal gray (PAG), as well as output axonal projections to laryngeal motoneurons, and to the respiratory rhythm generator in the preBötzinger complex.

5. These studies advance the important concept that the brainstem vocalization pattern generator integrates with the medullary respiratory pattern generator to control expiratory airflow as a key mechanism to produce various USV types characterized by different pitch patterns.

Weaknesses:

A limitation is that the cellular and circuit mechanisms by which the vocalization pattern generator integrates with the respiratory pattern generator to control expiratory airflow have not been fully worked out, requiring future studies.

Reviewer #2 (Public Review):

Summary:

Both human and non-human animals modulate the frequency of their vocalizations to communicate important information about context and internal state. While regulation of the size of the laryngeal opening is a well-established mechanism to regulate vocal pitch, the contribution of expiratory airflow to vocal pitch is less clear. To consider this question, this study first characterizes the relationship between the dominant frequency contours of adult mouse ultrasonic vocalizations (USVs) and expiratory airflow using whole-body plethysmography. Next, the authors build off of their previous work characterizing intermediate reticular oscillator (iRO) neurons in mouse pups to establish the existence of a genetically similar population of neurons in adults and show that artificial activation of iRO neurons elicits USV production in adults. Third, the authors examine the acoustic features of USV elicited by optogenetic activation of iRO and find that a majority of natural USV types (as defined by pitch contour) are elicited by iRO activation.

Strengths:

Strengths of the study include the novel consideration of expiratory airflow as a mechanism to regulate vocal pitch and the use of intersectional methods to identify and activate the iRO in adult mice. The establishment of iRO neurons as a brainstem population that regulates vocal production across development is an important finding.

Weaknesses:

The study does not include statistical analyses to compare the observed relationships between expiratory airflow and USV pitch to a null model in which expiratory airflow and USV pitch are unrelated. The findings of the study also do not provide clear evidence to support the authors' model in which distinct brainstem populations (iRO and RAm) independently regulate expiratory airflow and laryngeal adduction. Although this study establishes iRO as an important population that regulates USV production in adult mice, the question of whether and how different brainstem populations contribute differentially to vocal production remains an important open question. Lastly, the addition of statistical analyses would help to strengthen the study's conclusion that iRO activation positively biases the relationship between expiratory airflow and USV pitch across multiple USV types.

Author response:

The following is the authors’ response to the original reviews.

In the revised manuscript we have included an additional study that significantly contributes to the conclusions and models of the original version. Briefly, Figure 3 now describes our characterization of the diaphragm and laryngeal muscle activities (electromyography, EMG) during endogenous vocalizations. These EMGs also serve as representations of the brainstem breathing central pattern generator (CPG) inspiratory and post-inspiratory generating neurons, respectively. In our original submission, we found that many of the vocalizations had changes in pitch that mirrored the change in expiratory airflow (we termed positive intonation), and we proposed that the coordination of breathing muscles (like the inspiratory muscles) and larynx patterned this. This mechanism is akin to our findings for how neonatal cries are rhythmically timed and produced (Wei et al. 2022). The newly presented EMG data re-inforces this idea. We found that for vocalizations with positive intonation, the inspiratory diaphragm muscle has an ectopic burst(s) of activity during the expiration phase which corresponds to a decrease in airflow and pitch, and this is followed by laryngeal muscle activity and increased pitch. This can be cycled throughout the expiration to produce complex vocalizations with oscillations in pitch. A basal breath is hardwired for the laryngeal muscle activity to follow the diaphragm, so the re-cycling of this pattern nested within an expiration (a ‘mini-breath’ in a ‘breath’) demonstrates that the vocalization patterning system engages the entire breathing CPG. This contrasts with the canonical model that activity of the laryngeal premotor neurons control all aspects of producing / patterning vocalizations. Furthermore, this mechanism is exactly how the iRO produces and patterns neonatal vocalizations (Wei et al. 2022) and motivates the likely use of the iRO in adult vocalizations.

Response to recommendations for the authors:

Reviewer #1:

(1) The authors should note in the Discussion that the cellular and circuit mechanisms by which the vocalization pattern generator integrates with the respiratory pattern generator to control expiratory airflow have not been fully worked out, requiring future studies.

This was noted in the discussion section “The iRO likely patterns intonation for endogenous phonation”.

(2) Please change the labeling of the last supplemental figure to Figure Supplemental 5.

Thank you for identifying this.

Reviewer #2:

Major concerns

(1) While it is true that modulation of activity in RAm modulates the laryngeal opening, this statement is an incomplete summary of prior work. Previous studies (Hartmann et al., 2020; Zhang et al., 1992, 1995) found that activation of RAm elicits not just laryngeal adduction but also the production of vocal sounds, albeit vocal sounds that were spectrally dissimilar from speciestypical vocalizations. Moreover, a recent study/preprint that used an activity-dependent labeling approach in mice to optogenetically activate RAm neurons that were active during USV production found that re-activation of these neurons elicits USVs that are acoustically similar to natural USVs (Park et al., 2023). While the authors might not be required to cite that recent preprint (as it is not yet peer-reviewed), the fact that activation of RAm elicits vocal sounds is clear evidence that its effects go beyond modulating the size of the laryngeal opening, as this alone would not result in sound production (i.e., RAm activation must also recruit expiratory airflow). The authors should include these relevant studies in their Introduction. Moreover, the rationale for the model proposed by the authors (that RAm controls laryngeal opening whereas iRO controls expiratory airflow) is unclear with regard to these prior studies. The authors should include a discussion of how these prior findings are consistent with their model (as presented in the Introduction, as well as in Figure 4 and relevant Discussion) that RAm modulates the size of laryngeal opening but not expiratory airflow.

An introduction and discussion of the Veerakumar et. al. 2023 and Park et. al. 2024 manuscripts describing RAm in mice has now been included.

The iRO serves to coordinate the breath airflow and laryngeal adduction to produce sound and the intonation within it that mirrors the breath airflow. This occurs because the iRO can control the breathing CPG (synaptic input to the preBötC inspiratory pacemaker) and is premotor to multiple laryngeal muscles (Wei et. al. 2022). The modulation of the expiratory airflow is by inducing momentary contraction of the diaphragm (via excitation of the preBötC) which opposes (a.k.a. slows) expiration. This change in flow results in a decrease in pitch (Fig. 3 in the revised manuscript, Wei et. al. 2022).

It is our understanding that the basic model for RAm evoked USVs is that RAm evokes laryngeal adduction (and presumed abdominal expiratory muscle activation) and this activity is momentarily stopped during the breath inspiration by inhibition from the preBötC (Park et. al. 2024). So, in this basic model, any change in pitch and expiratory airflow would be controlled by tuning RAm activity (i.e., extent of laryngeal adduction). In this case, the iRO induced inspiratory muscle activity should not occur during expiration, which is not so (Fig. 3). Note, the activity of abdominal expiratory muscles during endogenous and RAm evoked USVs has not been characterized, so the contribution of active expiration remains uncertain. This is an important next step.

We have now included a discussion of this topic which emphasizes that iRO and RAm likely have reciprocal interactions (supported by the evidence of this anatomical structure). These interactions would explain why excitation of either group can evoke USVs and, perhaps, the extent that either group contributes to a USV explains how the pitch / airflow changes. An important future experiment will be to determine the sufficiency of each site in the absence of the other.

(2) The authors provide evidence that the relationship between expiratory airflow and USV pitch is variable (sometimes positive, sometimes negative, and sometimes not related). While the representative spectrograms clearly show examples of all three relationship types, no statistical analyses are included to evaluate whether the relationship between expiratory airflow and USV pitch is different than what one would expect by chance. For example, if USV pitch were actually unrelated to expiratory airflow, one might nonetheless expect spurious periods of positive and negative relationships. The lack of statistical analyses to explicitly compare the observed data to a null model makes it difficult to fully evaluate to what extent the evidence provided by the authors supports their claims.

We have now included two null distributions and compared our observed correlation values to these. The two distributions were created by taking each USV / airflow pair and randomly shuffling either the normalized USV pitch values (pitch shuffled) or the normalized airflow values (airflow shuffled) to simulate the distribution of data should no relationship exist between the USV pitch and airflow.

(3) The relationship between expiratory airflow and USV pitch comes with two important caveats that should be described in the manuscript. First, even in USV types with an overall positive relationship between expiratory airflow and pitch contour, the relationship appears to be relative rather than absolute. For example, in Fig. 2E, both the second and third portions of the illustrated two-step USV have a positive relationship (pitch goes down as expiratory airflow goes down). Nonetheless, the absolute pitch of the third portion of that USV is higher than the second portion, and yet the absolute expiratory airflow is lower. The authors should include an analysis or description of whether the relationship between expiratory airflow and USV pitch is relative vs.

absolute during periods of 'positive intonation'.

The relationship between pitch and airflow is relative and this in now clarified in the text. To determine this, we visualized the relationship between the two variables by scatterplot for each of the USVs syllables and, as the reviewer notes, a given airflow cannot predict the resulting frequency and vice versa.

(4) A second important caveat of the relationship between expiratory airflow and USV pitch is that changes in expiratory airflow do not appear to account for the pitch jumps that characterize mouse USVs (this lack of relationship also seems clear from the example shown in Fig. 2E). This caveat should also be stated explicitly.

The pitch jumps do not have a corresponding fluctuation in airflow, and this is now stated in the results and discussion.

(5) The authors report that the mode of relationship between expiratory airflow and USV pitch (positive intonation, negative intonation, or no relationship) can change within a single USV. Have the authors considered/analyzed whether the timing of such changes in the mode of relationship coincides with pitch jumps? Perhaps this isn’t the case, but consideration of the question would be a valuable addition to the manuscript.

We analyzed a subset of USVs with pitch jumps that were defined by a change >10 kHz, at least 5ms long, and had one or two jumps. The intonation relationships between the sub-syllables within a USV type were not stereotyped as evidenced by the same syllable being composed of combinations of both modes.

(6) The authors incorrectly state that PAG neurons important for USV production have been localized to the ventrolateral PAG. Tschida et al., 2019 report that PAG-USV neurons are located predominantly in the lateral PAG and to a lesser extent in the ventrolateral PAG (see Fig. 5A from that paper). The finding that iRO neurons receive input from VGlut2+ ventrolateral PAG neurons represents somewhat weak evidence that these neurons reside downstream of PAG-USV neurons. This claim would be strengthened by the inclusion of FOS staining (following USV production), to assess whether the Vglut+ ventrolateral PAG neurons that provide input to iRO are active in association with USV production.

This comment correctly critiques that our PAG à iRO tracing does not demonstrate that the labeled PAG neurons are sufficient nor necessary for vocalization. Directly demonstrating that activation and inhibition the PAG-iRO labeled neurons ectopically drives or prevents endogenous USVs is an important next step. While FOS implies this connectivity, it does not definitely establish it and so this experiment is impacted by some of the caveats of our tracing (e.g. PAG neurons that drive sniffing might be erroneously attributed to vocalization).

Our reading of the literature could not identify an exact anatomical location within the mouse PAG and this site appears to vary within a study and between independent studies (like within and between Tschida et. al. 2019 and Chen et. al. 2021). The labeling we observed aligns with some examples provided in these manuscripts and with the data reported for the retrograde tracing from RAm (Tschida et al 2019).

(7) In Figure S5A, the authors show that USVs are elicited by optogenetic activation of iRO neurons during periods of expiration. In that spectrogram, it also appears that vocalizations were elicited during inspiration. Are these the broadband vocalizations that the authors refer to in the Results? Regardless, if optogenetic activation of iRO neurons in some cases elicits vocalization both during inspiration and during expiration, this should be described and analyzed in the manuscript.

The sound observed on the spectrogram during inspiration is an artefact of laser evoked head movements that resulted in the fiber cable colliding with the plethysmography chamber. In fact, tapping an empty chamber yields the same broad band spectrogram signal. The evoked USV or harmonic band vocalization is distinct from this artefact and highlighted in pink.

(8) Related to the comment above, the authors mention briefly that iRO activation can elicit broadband vocalizations, but no details are provided. The authors should provide a more detailed account of this finding.

The broadband harmonic vocalizations we sometimes observe upon optogenetic stimulation of AAV-ChR2 expressing iRO neurons are akin to those previously described within the mouse vocal repertoire (see Grimsley et. al .2011). We have added this citation and mentioned this within the text.

(9) The effects of iRO stimulation differ in a couple of interesting ways from the effects of PAGUSV activation. Optogenetic activation of PAG-USV neurons was not found to entrain respiration or to alter the ongoing respiratory rate and instead resulted in the elicitation of USVs at times when laser stimulation overlapped with expiration. In contrast, iRO stimulation increases and entrains respiratory rate, increases expiratory and inspiratory airflow, and elicits USV production (and also potentially vocalization during inspiration, as queried in the comment above). It would be informative for the authors to add some discussion/interpretation of these differences.

We have added a section of discussion to describe the how these different results may be explained by the iRO being a vocal pattern generator versus the PAG as a ‘gating’ signal to turn on the medullary vocalization patterning system (iRO and RAm). See discussion section ‘The iRO likely patterns intonation for endogenous phonation’.

(10) The analysis shown in Fig. 4D is not sufficient to support the author’s conclusion that all USV types elicited by iRO activation are biased to have more positive relationships between pitch and expiratory airflow. The increase in the relative abundance of down fm USVs in the opto condition could account for the average increase in positive relationship when this relationship is considered across all USV types in a pooled fashion. The authors should consider whether each USV type exhibits a positive bias. Although such a comparison is shown visually in Fig. 4G, no statistics are provided. All 7 USV types elicited by optogenetic activation of iRO should be considered collectively in this analysis (rather than only the 5 types currently plotted in Fig. 4G).

In the original submission the statistical analysis of r values between opto and endogenous conditions was included in the figure legend (‘panels E-G, two-way ANOVA with Sidak’s post-hoc test for two-way comparisons was used; all p-values > 0.05), and this has not changed in the revised manuscript. We have now provided the suggested comparison of opto vs endogenous USVs without down fm (Fig. 5D). This positive shift in r is statistically significant (…).

(11) The evidence that supports the author’s model that iRO preferentially regulates airflow and that RAm preferentially regulates laryngeal adduction is unclear. The current study finds that activation of iRO increases expiratory (and inspiratory) airflow and also elicits USVs, which means that iRO activation must also recruit laryngeal adduction to some extent. As the authors hypothesize, this could be achieved by recruitment of RAm through iRO’s axonal projections to that region.

Note, it is more likely that iRO is directly recruiting laryngeal adduction as they are premotor to multiple laryngeal muscles like the thyroarytenoid and cricothyroid (Wei et. al. 2022). The ‘Discussion’ now includes our ideas for how the iRO and RAm likely interact to produce vocalizations.

In the recent preprint from Fan Wang’s group (Park et al., 2023), those authors report that RAm is required for USV production in adults, and that activation of RAm elicits USVs that appear species-typical in their acoustic features and elicits laryngeal adduction (assessed directly via camera). Because RAm activation elicits USVs, though, it must by definition also recruits expiratory airflow. Can the authors add additional clarification of how the evidence at hand supports this distinction in function for iRO vs RAm?

See response to ‘Major Concern #1”.

Minor concerns

(1) The authors might consider modifying the manuscript title. At present, it primarily reflects the experiments in Figure 2.

We have provided a title that we feel best reflects the major point of the manuscript. We hope that this simplicity enables it to be recognized by a broad audience of neuroscientists as well as specialists in vocalization and language.

(2) The statement in the abstract that "patterns of pitch are used to create distinct 'words' is somewhat unclear. Distinct words are by and large defined by combinations of distinct phonemes. Are the authors referring to the use of "tonemes" in tonal languages? If so, a bit more explanation could be added to clarify this idea. This minor concern includes both the Abstract, as well as the first paragraph of the Introduction.

We have clarified this line in the abstract to avoid the confusing comparison between mouse vocalizations and human speech. In the introduction we have expanded our explanation to clarify that variations in pitch are a component of spoken language that add additional meaning and depth to the underlying, phonemic structure.

(3) Multiple terms are used throughout the manuscript to refer to expiratory airflow: breath shape (in the title), breath pattern, deviations in exhalation, power of exhalation, exhalation strength, etc. Some of these terms are vague in meaning, and a consolidation of the language would improve the readability of the abstract and introduction.

We have chosen a smaller selection of descriptive words to use when describing these breath features.

(4) Similarly, "exhalation" and "expiration" are both used, and a consistent use of one term would help readability.

See point 3.

(5) In a couple of places in the manuscript, the authors seem to state that RAm contains both laryngeal premotor neurons as well as laryngeal motor neurons. This is not correct to our knowledge., but if we are mistaken, we would ask that the authors add the relevant references that report this finding.

It is our understanding that the RAm is defined as the anatomical region consistent with the murine rostral and caudal ventral respiratory groups composed of multiple premotor neuron pools to inspiratory, expiratory, laryngeal, and other orofacial muscles. This is supported by neurons within RAm that reflect multiple phases of the inspiratory and expiratory cycle (Subramanian et. al. 2018) and excitation of sub-regions within RAm modulating multiple parts of the breathing control system (Subramanian et. al. 2018 and Subramanian 2009). Rabies tracing of the various premotor neurons which define the anatomical region of RAm in the mouse shows that they surround the motor neurons in the loose region of the nucleus ambiguus (the anatomical location of RAm) for multiple muscles of the upper airway system, such as the thyroarytenoid (Wu et. al. 2017, Dempsey et. al. 2021 and Wei et. al. 2022). Given that the name RAm reflects a broad anatomical location, we have used it to describe both the premotor and motor neurons embedded within it. We have now clarified this in the text.

(6) The statistical analysis applied in Figure 1C is somewhat confusing. The authors show two distributions that appear different but report a p-value of 0.98. Was the analysis performed on the mean value of the distributions for each animal, the median, etc.? If each animal has two values (one for USV+ breaths and one for USV- breaths), why not instead compare those with a paired t-test (or Wilcoxon rank sign)? Additional information is needed to understand how this analysis was performed.

The original manuscript version used a two-way anova to compare the normalized histogram of instantaneous frequency for breaths with (USV+) or without (USV-) for each animal (first factor: USV+/-, second factor: Frequency). The p-value for the first factor (USV) was 0.98 showing no statistically significant effect of USV on the distribution of the histogram.

For simplicity, we have instead performed the analysis as suggested and include a bar graph. This analysis shows that the instantaneous frequency of USV breaths is, in fact, statistically significantly lower than those without USVs. We have updated the figure legend and text to reflect this.

(7) The use of the word "syllable" to describe parts of a USV that are produced on a single breath may be confusing to some scientists working on rodent USVs. The term 'syllable' is typically used to describe the entirety of a USV, and the authors appear to use the term to describe parts of a USV that are separated by pitch jumps. The authors might consider calling these parts of USVs "sub-syllables".

We have clarified these descriptions throughout the text. We now refer to the categories as ‘syllable types’, define ‘syllables’ as ‘a continuous USV event’ with no more than 20ms of silence within and finally ‘sub-syllables’ to refer to components of the syllable separated by jumps in frequency (but not gaps in time).

(8) In Figure S3, final row, the authors show a USV produced on a single breath that contains two components separated by a silent period. This type of bi-syllabic USV may be rare in adults and is similar to what the authors showed in their previous work in pups (multiple USVs produced on a single expiration, separated by mini-inspirations). One might assume that the appearance of such USVs in pups and their later reduction in frequency represents a maturation of vocalrespiratory coordination. Nonetheless, the appearance of bi-syllabic USVs has not been reported in adult mice to our knowledge, and the authors might consider further highlighting this finding.

We were also struck by the similarity of these USVs to our study in neonates and such types of similarities sparked an interest in the role of the iRO in patterning adult USVs. We now include a description of the presence and abundance of bi- and tri-syllablic calls observed in our recordings to highlight this finding.

(9) Figure 4 is referenced at the end of the second Results section, but it would seem that the authors intended to reference Figure 2.

For simplicity we included some of the referenced data within Fig. S5. We appreciate the recommendation.

(10) In the optogenetic stimulation experiments, the authors should clarify why bilateral stimulation was applied. Was unilateral stimulation ineffective or less effective? The rationale provided for the use of bilateral stimulation (to further localize neural activation) is unclear.

The iRO is bilateral and, we presume, functions similarly. So, we attempted to maximally stimulate the system. We have clarified this in the methods.

(11) Figure Supplemental '6' should be '5'.

Thanks!

(12) Last sentence of the Introduction: "Lasty" should be "lastly".

Thanks!

(13) There are two references for Hage et al., 2009. These should be distinguished as 2009a and 2009b for clarity.

Thanks!

Version 2

Abstract

Abstract

Intonation in speech is the control of vocal pitch to layer expressive meaning to communication, like increasing pitch to indicate a question. Also, stereotyped patterns of pitch are used to create distinct sounds with different denotations, like in tonal languages and, perhaps, the ten sounds in the murine lexicon. A basic tone is created by exhalation through a constricted laryngeal voice box, and it is thought that more complex utterances are produced solely by dynamic changes in laryngeal tension. But perhaps, the shifting pitch also results from altering the swiftness of exhalation. Consistent with the latter model, we describe that intonation in most vocalization types follows deviations in exhalation that appear to be generated by the re-activation of the cardinal breathing muscle for inspiration. We also show that the brainstem vocalization central pattern generator, the iRO, can create this breath pattern. Consequently, ectopic activation of the iRO not only induces phonation, but also the pitch patterns that compose most of the vocalizations in the murine lexicon. These results reveal a novel brainstem mechanism for intonation.

Introduction

Modulation of the frequency of produced sound, perceived as pitch, creates meaning within words or phrases through intonation (Prieto et al., 2015). For example, in English, an increasing pitch is used to indicate a question or stress importance and a decreasing pitch communicates a declaration. Additionally, the concatenation of specialized sounds accented by variations in pitch enriches the composition of spoken language, like how the same sound at divergent pitches can relay different meanings (Howie et al., 1976). Two key pieces of the phonation system are the larynx (the “voice-box”) and the breathing muscles (Berke et al., 2009, Laplange et al., 2018). Succinctly, the breathing muscles drive airflow through a narrowed larynx to produce a basic vocalization (Finck et al., 2009). The local speed of the airflow through the larynx dictates the fundamental frequency of the tone, so changes in either the swiftness of the breath exhalation or the extent of laryngeal closure can both, presumably, alter the pitch (Kelm-Nelson et al., 2018, Herbst C.T., 2016, Mahrt et al., 2016). While control of the size of the laryngeal opening is well established as a mechanism to regulate the dynamic changes in pitch for human to rat and mouse vocalizations (Poeppel et al., 2020, Johnson et al., 2010, Riede et al., 2017), the contribution of exhalation itself remains to be carefully defined. In fact, it is presumed that the velocity of expiration only modulates the vocal amplitude or loudness (Riede T., 2011, Riede T., 2013). This perception stems from the airflow of the rodent breath not strongly predicting the pitch. Yet paradoxically, an injection of air below the larynx to enhance flow increases pitch (Riede T., 2011). This incongruity even extends to songbirds, a leading vocalization model system (Suthers et al., 2002, Schmidt et al., 2014, Plummer et al., 2008, Goller et al., 2004). Here, we seek to resolve this inconsistency by taking advantage of the experimental, behavioral, and genetic approaches in the mouse (Yackle K., 2023). If two independent variables are used to alter pitch, like the larynx and the breath airflow, then the interplay would enhance the ability to produce a diverse repertoire of sounds and thereby enable a broader lexicon.

The medullary brainstem possesses at least two means that might account for the two control points proposed above, laryngeal diameter and exhalation speed. First, modulation of some laryngeal premotor neurons in the retroambiguus (RAm) and perhaps intermingled or nearby motor neurons modulates the size of the laryngeal opening (Kelm-Nelson et al., 2018, Hage S.R., 2009a, Park et al., 2024) and is even sufficient to evoke vocalizations across species, albeit mostly abnormal (Zhang et al., 1995, Hartmann & Brecht, 2020, Veerakumar et al., 2023, Park et al., 2024). Second, the vocalization central pattern generator we recently described, called the intermediate Reticular Oscillator (iRO), induces coordinated changes in the expiratory airflow and laryngeal closure (Wei et al., 2022). For example, during neonatal cries, the iRO oscillates exhalation speed and larynx activity to time the syllable sounds. Thus, the RAm provides a mechanism to modulate pitch by controlling laryngeal diameter independently from the iRO altering the tone by dictating the speed of the breath expiratory airflow. While the contribution of Ram in adult phonation has recently been established in mice (Veerakumar et al., 2023, Park et al., 2024), the role of the iRO remains undefined.

Here, we describe the coordinated changes in breath airflow and pitch in the ten vocalizations of the adult murine lexicon (Grimsley et al., 2011). We describe that the modulation of pitch for the different vocalizations either correlates or anticorrelates with the changes in exhalation velocity. These results support a model in which two independent mechanisms involving changes in laryngeal opening airflow control tone. Consistently, we found that the pattern of breathing muscle activity is different for each mechanism. For example, the muscle that drives the inspiration phase of the breath is ectopically re-engaged to break expiratory airflow and decrease pitch. This mirrors how the iRO rhythmically patterns the breathing muscles to time syllables during neonatal cries. Using anatomical, molecular, and functional approaches, we demonstrate that the iRO vocal central pattern generator drives changes in breath expiratory airflow to pattern pitch and produce seven of the ten vocalizations in the endogenous lexicon. These data resolve the prior paradoxical role of exhalation speed in sound production and show it can directly control pitch. Additionally, we establish the iRO as a mechanistic basis for intonation. And lastly, these results generalize the crucial role of the iRO in phonation across developmental stages and, we presume, across species.

Results

Vocalizations are produced by a program coupled to breathing

It is possible that the ten murine ultrasonic vocalizations (USVs) defined by unique pitch patterns (‘syllable types’; Grimsley et al., 2011) are formed by distinct breaths or as substructures nested in a common breath. Prior work has suggested the latter (Sirotin et al., 2014). To expand upon this, we simultaneously measured breathing and USVs by customizing the lid of a whole-body plethysmography chamber to accommodate a microphone. Male mice in the chamber were exposed to fresh female urine and robustly sniffed and vocalized for the first 5-10 minutes of the recording at a peak rate of about 4 events per second (n = 6) (Fig. 1A-B). A vocalization was classified as a narrow-band sound in the 40-120 kHz ultrasonic frequency range during a single breath (Fig. 1A). The instantaneous frequency of vocalization breaths was typically between 5-10 Hz (mean: 7.5 Hz) (Fig. 1C) and mostly occurred during episodes of rapid sniffs (8.5 - 10 Hz) (Fig. 1A and B), as previously reported (Sirotin et al., 2014, Castellucci et al., 2018). When compared to neighboring breaths, vocalization breaths were slightly slower overall (Fig. 1C) with subtly larger inspiratory and perhaps expiratory airflow despite similar durations of each phase (Fig. 1D and E). These data reveal that a vocalization breath appears mostly like a normal breath but with the addition of a nested sound pattern. This led us to hypothesize that a distinct sub-program is activated within a breath to generate a vocalization.

The full repertoire of vocalizations occurs within a normal appearing breath.

A, Male mice exposed to female urine produce ultrasonic vocalizations (USV) at about 75 kilohertz (top) that coincide with the expiratory airflow (E, arbitrary units) of the breath cycle (bottom). Red box indicates the length of the USV. A bout of vocalizations contains breaths with USVs (red) interspersed among sniff breaths (black). B, Rates of breathing (black) and USV production (red). Exposure to female urine at time 0, n=6 mice. C, Left, histogram of the instantaneous frequency of breaths with and without USVs from n=6 animals. Right, average instantaneous frequency for each mouse (mean ±SEM). Each dot is the mean from each animal. p-value 0.03; paired t-test. D, Scatter plot of the inspiratory (Ti) and expiratory time (Te) for USV (red) and non-USV (black) breaths from a representative animal. Right, bar graph of mean ± SEM of Ti, Te, and the ratio for n=6. p-values 0.40, 0.18, and 0.25; paired t-test. E, The breath peak inspiratory (pif) and expiratory (pef) airflow represented as in D. p-values 0.01, 0.27; paired t-test. F, Bar graph (mean ±SEM) of the percent of total USVs for each type from n=6 mice.

The adult murine lexicon is composed of at least 10 USV syllable types that are defined by different, but stereotyped, patterns of pitch. Most breaths contain a single syllable (88%), which we define as a continuous USV event (Fig. 1F and S2-3), however, on some occasions, we observed 2 (8%) or 3 (4%) syllables separated by >20 ms within a single breath (Fig. S3). This structure mirrors that described in neonatal cries (Wei et al., 2022). A pre-trained convolutional neural network (CNN) was used to classify USVs into different types based on changes in pitch (Fonseca et al., 2021), and the on- and offset of each vocalization was overlayed upon the corresponding breath airflow (Fig. S1-3). Vocalizations began and ended throughout expiration (Fig. S1), and the most common tended to start near the onset of exhalation and ended shortly there-after (like the up frequency modulated, step down, flat, and short types) (Fig. S2). Vocalizations with more intricate changes in pitch had more variable times of on- and offset (like complex, chevron, two step, multi, step up, down frequency modulated) (Fig. S3). And lastly, when the vocalizations occurred late in expiration, the duration of this breath phase was prolonged (Fig. S4). The bias of USV timing by the breath, combined with the USV modulation of breath length demonstrate these programs are independent but reciprocally coupled.

Two mechanisms create the changes in pitch pattern

Fluctuations in airflow speed through the larynx produce changes in the sound’s pitch. For example, augmenting airflow through the explanted rodent larynx increases pitch (Mahrt et al., 2016). We proposed two potential mechanisms that could contribute to how the laryngeal airflow is modulated to form the distinct USV types in the murine lexicon: one based on the swiftness of exhalation pushing air through the larynx (model 1), and the other based on the diameter of the laryngeal opening (model 2). According to the first model, changes in pitch positively correlate with the breath expiratory airflow measured by plethysmography, which we term positive intonation. On the other hand (model 2), a narrowed larynx increases pitch by speeding local airflow while simultaneously impeding the overall expiratory airflow measured by plethysmography; we call this negative intonation. Note, these models can form similar expiratory airflow patterns, but predict opposite relationships to pitch.

We assessed for evidence of each model by calculating the correlation coefficient (r) between instantaneous expiratory airflow and the corresponding USV fundamental frequency. Down or up frequency modulated (fm) USVs served as simple USV examples and we found that these were positively or negatively correlated, respectively (median r = 0.62 and −0.46, Fig. 2A and C). These two were also distinguished by the sound onset and offset, whereby down fm started and ended later during the expiration (Fig. S2-3). These distinguishing features are consistent with the sounds being produced by separate mechanisms to alter pitch, positive and negative intonation.

The ten types of USVs are produced by at least two mechanisms that modulate airflow.

A, Left, example of the expiratory airflow and pitch for a down frequency modulated (fm) USV. Middle, magnification of airflow and sound. The scale of airflow is not displayed. The time of breath airflow from expiration onset during the USV is color coded blue to white. Scatter plot of instantaneous expiratory airflow and pitch for the single USV and the correlation (line, r). Note, the change in pitch mirrors airflow (annotated as “+”), consistent with the ‘breath modulation’ called model 1. Box and whisker plot of n=40 down fm correlation coefficients (r). Controls, normal USV pitch vs shuffled airflow or shuffled pitch to normal USV expiratory airflow. B-D, Representative expiratory airflow and pitch, box and whisker plot of all r values, and onset / offset for complex (n=165), up fm (n = 589), and two step (n=61) vocalizations represented as in A. The airflow for each unique USV element is uniquely color coded as green, blue, or purple. Note, the change in pitch for two components correlates and one anticorrelates. This is consistent with both mechanisms being sequentially used. Annotated as mixed blue and green box and whisker plot. * = p < 0.05, one-way ANOVA with Sidak’s post-hoc test.

Between these mechanisms, most syllable types displayed positive intonation. Five of the other eight USV syllable types had positively shifted intonation, the complex, step down, chevron, two step, and multi (median r = 0.31, 0.28, 0.32, 0.19, and 0.24 respectively) (Fig. 2B, D and Fig. S5). In particular, the mirrored oscillations in the breath expiratory airflow and pitch during a complex vocalization best illustrated the positively coupled relationship (model 1, Fig. 2C). In contrast, the up fm was the only USV with negative intonation. The r values for positively and negatively correlated USVs were not random, as they fell outside r values computed in simulated datasets composed of USVs with shuffled expiration airflow or expirations with shuffled pitch frequencies (Fig. 2 and Fig. S5).

The two step and step up USVs (median r = 0.19 and −0.03) appeared to have a portion of the pitch pattern correlated with the expiratory airflow, while the other part(s) were un- or anti-correlated (e.g., the two step, Fig. 2E). This suggests that the pitch is produced by switching between positive and negative intonation mechanisms within the breath. The remaining two USV types (flat and short) occurred with various breath shapes which resulted in a wide range of r values (Fig. S5). Some USV syllable types are defined by the presence of large, instantaneous jumps in frequency (like the step up, step down, two- and multi-step). A jump was not associated with a corresponding change in airflow. Also, the jumps did not predict the subsequent intonation pattern. For example, step up syllables could be negative to negative (20%), negative to positive (20%), or positive to positive (60%) intonations.

Across all USV types, we found that the relationship between airflow and pitch is relative rather than absolute which is expected since the relationship is determined by at least two independent variables (laryngeal tone and exhalation speed). In summary, these results establish an important positive connection between the breath expiratory airflow to modulate pitch (model 1). This supports the hypothesis that a vocalization pattern generator must integrate with and even control the breath airflow as a key mechanism to produce various USV types in the murine lexicon.

Inspiratory and laryngeal muscles have coordinated activity that represents positive and negative intonation

To explore the mechanisms underlying the two intonation models, we simultaneously recorded the electrical activity of the primary muscles for breathing, the diaphragm, and laryngeal control (thyroarytenoid and cricothyroid) during basal breathing and bouts of vocalizations in male mice. Electromyography (EMG) electrodes were permanently placed along the diaphragm and inserted into the larynx, while breathing and USVs were measured in parallel by whole-body plethysmography and a microphone (as in Fig. 1). Electrocardiogram signals were annotated and removed from the diaphragm EMG recordings post-hoc. According to our models and the timing of the syllable types within the breath (Fig. S2-3), we anticipated that USVs with positive intonation would have a coordinated re-activation of the diaphragm and laryngeal muscles later in expiration, while the up fm would only have laryngeal activity at expiration onset.

EMG activity during basal breathing displayed the three phases of the breath cycle (Yackle, 2023). The diaphragm was active during inspiration and larynx in the subsequent post-inspiration period, and these were followed by a late expiration phase where neither muscle was active (Fig. 3A). The airflow measured by plethysmography had a ∼10ms lag when compared to the EMG activities (blue and red arrows, Fig. 3A). Upon exposure to female odors, the male mice produced bouts of USVs containing the entire lexicon for several minutes, demonstrating that electrodes interfered with neither breathing nor sound production (Fig. 3B). Breaths containing USVs were distinguished from adjacent breaths lacking sound by an increase in laryngeal muscle activity that prolonged into expiration. Like airflow, the sound followed EMG activity by ∼10ms. More than 10 mice were studied and, unsurprisingly, given the invasiveness of the EMGs only three produced robust signals and the number of acquired vocalizations varied from tens to thousands between these mice (n=70, 1482, 2819). The analysis below is from a single animal that had the clearest EMG signals, but the same results and relationships were observed in all three animals.

Ectopic activation of inspiratory and laryngeal muscles corresponds to changes in vocalization pitch.

A, Activity of the diaphragm (inspiratory) and laryngeal (thyroarytenoid and cricothyroid) muscles were recorded in vivo by electromyography (EMG) simultaneously with breathing and USVs. Right, example of sound, breath airflow, and muscle activities during a basal breath. The diaphragm shows restricted activity during inspiration and laryngeal muscles are active during the post-inspiration period. Note, blue and red arrows / lines indicate the ∼10ms offset of the peak EMG activity and airflow / sound measurements. B, Representative vocalization bout shows robust vocalizations and breathing in mice with implanted EMGs. C, Representative sound, airflow, diaphragm and laryngeal EMGs during the expiration of a down fm (n = 23 annotated), complex (n = 43 annotated), and up fm USV (n = 29 annotated). Blue arrow and dashed line indicate the airflow and sound offset from the diaphragm peak muscle activity. Bottom, probability density function (PDF) for the peak of the integrated EMG activity for the diaphragm (blue) and larynx (red) during the normalized expiration. Y – axis is from 0 to 1. Note, data in A-D is from one animal with clear EMG signals that represents the findings in all three animals studied. D, Schematics for each model. Model 1 – breath control: inspiratory and laryngeal muscles have alternating activity throughout the sound / expiration and a r > 0 for pitch vs expiratory airflow. Muscle activities correspond to an increase (laryngeal) and a decrease (diaphragm) in pitch. Model 2 – laryngeal only: laryngeal but not diaphragm activity occurs during the sound and produces a r < 0 for pitch vs expiratory airflow.

To study positive intonation, we analyzed the down fm and complex USVs. Down fms started later during expiration and began at a drop in expiratory airflow (Fig. 3C). The diaphragm activity occurred just before the decrease in expiratory airflow (blue arrow, Fig. 3C) and the following laryngeal activity persisted during the sound. This pattern reflects the activity of the muscles during a normal breath, inspiration (diaphragm) then post-inspiration (larynx), but this pattern was ectopically embedded within an expiration. Similarly, complex vocalizations had the diaphragm then laryngeal activity pattern during expiration, but this cycled multiple times concurrent with the increase and decrease in expiratory airflow and pitch (Fig. 3C). Complex USVs had an average of 1.5 cycles per expiration, the interval between diaphragm bursts was 43 ± 14 ms, and the laryngeal activity occurred at 69 ± 12% through this diaphragm-to-diaphragm interval. Also, in ∼19% of the complex USVs, sound, laryngeal, and diaphragm activity co-occurred, suggesting that other mechanisms contribute to the diversity of sounds.

Up fm represents negative intonation, and correspondingly, the activity of the muscles distinct from the positively intonated types. At the peak of the post-inspiratory period of the breath, the vocalization began and persisted throughout the laryngeal muscle activity (Fig. 3C). The sound ended at the onset of a burst in the diaphragm EMG (Fig. 3C).

In summary, these EMG studies of the key inspiratory muscle and larynx serve to reflect the core components of the breathing central pattern generator (CPG) that produce inspiration and post-inspiration. Ectopic activation of these antiphase patterns during the expiration of a vocal breath appears to result in a USV with positive intonation and the cyclic engagement of this leads to an oscillating pitch. Additionally, the termination of the USV with negative intonation corresponds to the re-activation of inspiratory muscles (Fig. 3D). The novel finding that the endogenous pattern of the breathing CPG is re-engaged within an adult vocal breath, a ‘mini-breath’, mimics the rhythmic syllables of the neonatal vocalizations patterned by the intermediate Reticular Oscillator (iRO;Wei et al., 2022). This suggests that the iRO also plays a central role in the production of adult vocalizations.

The iRO resides within the adult brainstem phonation circuit

The iRO has yet to be identified in adult mice. The iRO is molecularly defined in the neonate by the co-expression of Preproenkephalin (Penk) and Vesicular glutamate transporter 2 (Vglut2) and is anatomically localized to the medullary ventral intermediate Reticular Formation (iRT) directly medial to the compact nucleus ambiguus (Wei et al., 2022). We determined that the iRO molecular and anatomical features exist in adults in two ways. First, we generated triple transgenic mice that label Penk+Vglut2+ neurons and the derived lineages with tdTomato (Penk-Cre; Vglut2-Flp; Ai65) (Fig. 4A). And second, we stereotaxically injected the iRO region of Penk-Cre; Vglut2-Flp mice with a Cre and Flp dependent reporter adeno-associated virus (AAV CreONFlpON-ChR2::YFP) (Fig. 4B-C). Consistent with the definition of the iRO in neonatal mice, tdTomato+ and YFP+ Penk+Vglut2+ neurons were found in the iRT adjacent to the compact nucleus ambiguus (Fig. 4A-C). These results demonstrate that the ventrolateral medulla of adult mice contains neurons with the molecular and anatomical identity of the iRO.

Anatomically and molecularly defined iRO neurons form a brainstem phonatory circuit.

A, Labeling of Penk+Vglut2+ neurons in the iRO anatomical region in adult Penk- Cre;Vglut2-Flp;Ai65 mice (CreONFlpON-tdTomato) (observed in n=5 mice). The iRO region is defined as medial to the compact nucleus ambiguus (cNA, ChAT +) in the ventral intermediate Reticular Tract (iRT). Note, the cNA is filled with tdTomato labeled axons. Cell bodies marked with arrowhead. B, Bilateral stereotaxic injection of AAV CreONFlpON-ChR2::EYFP into the iRO anatomical region of Penk-Cre;Vglut2-Flp adult mice. C, Magnified boxed region in B. Arrowheads label neuron soma quantified right (n=3). D, Axons of EYFP expressing iRO neurons from B in the retroambiguus (RAm) anatomical region where laryngeal premotor and motor neurons are located. E, Axons of EYFP expressing iRO neurons from B in the breathing pacemaker. F, Unilateral retrograde AAV CreON-EYP (AAVrg) stereotaxic injection into the iRO region in Vglut2- Cre adults (n=3). Glutamatergic neurons were identified in the contralateral (contra.) and ipsilateral (ipsi.) midbrain periaqueductal gray (PAG). Anatomical regions of the PAG: dorsomedial (dm), dorsolateral (dl), lateral (l), ventrolateral (vl) nearby to the dorsal raphe nucleus (DRN) and surrounding the cerebral aqueduct (Aq). Quantification of glutamatergic PAG neurons in each region demarcated, ns = not statistically significant; two-way ANOVA with Sidka’s post-hoc test. G, Model schematic of the iRO as a central component of the brainstem phonation circuit to convert a vocalization “go” cue from the PAG into a motor pattern.

Neonatal iRO neurons are presynaptic to the kernel of breathing, the pacemaker for inspiration (preBötzinger Complex, preBötC) (Smith et al., 1991), and premotor to multiple laryngeal and tongue muscles. We traced the YFP+ axons of Penk+Vglut2+ neurons (Penk-Cre; Vglut2-Flp and AAV CreONFlpON-ChR2::YFP) and found they elaborated within the nucleus ambiguus (NA) and retroambiguus (RAm) where laryngeal premotor and motor neurons localize (Fig. 4A, D), the breathing pacemaker (Fig. 4E), and the hypoglossal (tongue) motor nucleus (Fig. 4E). The projection patterns of these Penk+Vglut2+ neurons provide additional evidence that these adult neurons maintain the same connectivity properties as the neonatal iRO neurons, indicating they can control the key elements for vocalization: the breath airflow and larynx.

In adult mice, vocalizations have been triggered by activation of the midbrain periaqueductal gray (PAG), namely glutamatergic neurons in the lateral to ventrolateral subregion (Michael et al., 2020, Chen et al., 2021, Tschida et al., 2019). Note, an exact definition of PAG-USV stimulating neurons remains to be satisfactorily described. To assess if the iRO region is positioned downstream of the ventrolateral PAG, we unilaterally injected Vglut2-Cre mice with a CreON-ChR2::YFP expressing retrograde traveling AAV (AAVrg) (Fig. 4F). Among the labeled brain regions, we found YFP+ neurons in a region of the midbrain PAG overlapping with areas that contain PAG-USV neurons. To our surprise, neurons from the ipsi- and contralateral PAG projected to the iRO region in nearly equal numbers (Fig. 4F). These molecular, anatomical, and neural morphology characterizations reveal that the iRO exists in adults and is embedded within the brainstem phonation network (PAG ➔ iRO ➔ the preBötC, NA, RAm, hypoglossal) (Fig. 4G).

Ectopic activation of the putative iRO induced vocalization

If these labeled Penk+Vglut2+ neurons are indeed the iRO, we anticipated that ectopic activation would induce vocalization. We tested this in two ways. First, we generated Penk-Cre;Vglut2-Flp;CreONFlpON-ReaChR triple transgenic mice which express the red-shifted Channel Rhodopsin in Penk+;Vglut2+ neurons and the derived lineage (ReaChR mice) and second, we stereotaxically injected the AAV CreONFlpON-Channel Rhodopsin2::YFP (ChR2) into the iRO region of Penk- Cre;Vglut2-Flp mice. In both instances we implanted optic fibers above the iRO bilaterally to further localize neural activation (Fig. 5A and S6A). In both experimental regimes, ectopic light activation of the Penk+Vglut2+ neurons induced bouts of vocalizations where the breathing rate was entrained by the frequency of stimulation (Fig. 5A, S6A, S6I). Most bouts and the breaths within contained vocalizations (Fig. 5B and S6B), and the amplitudes of all elicited breaths were significantly increased (Fig. S6J). Some AAV-ChR2 mice showed previously described broad band harmonic vocalizations (like Grimsley et al., 2011), while others did not vocalize (n=5/9), likely due to incomplete labeling. Note, the audible sound during inspirations in these animals reflects orofacial movement artifacts that result in the fiberoptic colliding with the plethysmography chamber walls. Additionally, the ReaChR animals without vocalizations were found to have “off target” optic fiber implants (n=2/6). Taken together, these data are consistent with the notion that the iRO is sufficient to induce phonation via control of both breath airflow and laryngeal opening, just as it does in neonatal cries.

Ectopic activation of the iRO evokes airflow correlated USV types and switches the relationship of the anti-correlated types.

A, Optogenetic activation of the iRO region in Penk-Cre;Vglut2-Flp;CreONFlpON-ReaChR mice evokes USVs (blue box, 5 Hz stimulation). USVs occur during, or shortly after laser onset. B, Percentage of stimulation bouts containing at least one USV and the percentage of breaths within the stimulation window containing a USV. C, Percentage of optogenetically evoked (blue) or endogenously occurring (gray) syllables that are up fm of down fm. **** p-value< 0.001; two-way ANOVA with Sidak’s post-hoc test, p > 0.05 for all other types. D, Box and whisker plot of the correlation coefficient between breathing airflow and pitch (r) for all opto evoked (n=395) and endogenous (n=1850) USVs and all opto evoked and endogenous vocalizations without down fm (n=143 and 1810) from n=4 opto and n=6 endogenous mice. **** p-value < 0.001; Mann-Whitney test. E, Left, example of the expiratory airflow and pitch for an optogenetically evoked up fm USV. Middle, magnification of airflow and sound. Time of breath airflow during the USV is color coded blue to white. Scatter plot of instantaneous expiratory airflow and pitch for the single USV. Compare to endogenous up fm USV in Fig. 2D. Right, box and whisker plot of correlation coefficients (r) for each optogenetically evoked and endogenous up fm USV (n=15 vs. 589). F, Step up USV as in E. box and whisker plot of correlation coefficients (r) for each optogenetically evoked and endogenous step up USV (n=15 vs. 40). G, r value box and whisker plots for the remaining optogenetically evoked USV types (down fm n=242 vs. 40, step down n=38 vs. 293, flat n=34 vs. 337, short n=27 vs. 168, and chevron n=10 vs. 99 from n=4 opto and 6 endogenous mice. E-G, Two-way ANOVA with Sidak’s post-hoc test for two way comparisons was used; all p-values >0.05. H, Schematic illustrating the two mechanisms to pattern the USV pitch. Left, the reciprocal connection between the iRO and breathing pacemaker patterns the USVs with a positive correlation between pitch and airflow. Right, the retroambiguus (RAm) control of the larynx dictates the anti-correlated USV types. Middle, the combination of these two mechanisms within a single breath create additional USV patterns.

To demonstrate the specialization of the iRO neurons for vocalization and the inability of modulated breathing alone to elicit USVs, we performed several additional control experiments. First, to ensure that just stimulation of breathing is insufficient to elicit vocalization, we optogenetically excited the glutamatergic preBötC neurons (Vglut2-Cre with AAV CreON-ChR2). Indeed, we found that, although breathing sped up, optogenetic stimulation never elicited vocalizations (Fig. S6C, H, I). And second, to determine if the ability to elicit vocalizations was generalizable to other neural types in the iRO anatomical region, we activated Penk+, µ-opioid receptor+Vglut2+, Tachykinin 1+, and Vesicular GABA transporter+ neurons and found that vocalizations were never induced upon light stimulation, although breathing was altered in various ways (Fig. S6). In summary, these data functionally demonstrate the existence of Penk+Vglut2+ iRO neurons in adult mice and their ability to create vocalizations by modulating both breathing and presumably the larynx.

Excitation of the iRO evoked nearly the entire murine lexicon

Above, we described that one mechanism for generating the different patterns of vocalizations was via the modulation of the breath airflow (positive intonation). Once again, this was defined as a positive correlation between expiratory airflow and pitch (Fig. 2). We hypothesized that this property stems from the iRO’s capacity to control breathing, and so we made the following predictions: 1) that the USVs evoked after stimulation would be biased to those with an endogenous positive correlation between airflow and pitch (like the down fm and step down), and 2) that the elicited USVs would be transformed to become more positively correlated.

We classified the evoked iRO vocalizations (Penk-Cre;Vglut2-Flp;CreONFlpON-ReaChR) with the CNN, and to our surprise, seven of the ten types of endogenous USVs were induced upon activation of the iRO (Fig. 5). The most abundant elicited USV was the down fm which, in the endogenous dataset, had the strongest positive intonation (Fig. 5G and Fig. 2). Conversely, the USV with the strongest negative intonation was rarely found, up fm. These results are striking since the down fm is the least common endogenous USV and up fm is the most common (Fig. 2). These results are consistent with the first prediction where the optically evoked USV types were biased towards those with endogenous positive intonation. Beyond this, all the ectopic USVs combined had a more positive association between airflow and pitch compared to all endogenous USVs (Fig. 5D). This was not explained purely by the abundance of down fm since, when these were omitted from the analysis the positive bias was unchanged. This aligns with the second prediction. These data demonstrate that the iRO is sufficient to pattern nearly all USV types, and that the pitch of the induced vocalizations tightly follows the breathing airflow.

Discussion

Here we propose that the intonation that establishes the diversity of the adult murine lexicon is explained by two mechanisms, the modulation of the breath waveform and presumably the size of the laryngeal opening. We describe that unique vocalization types have characteristic fluctuations in the expiratory airflow, whereby some changes in pitch are strongly correlated with airflow (positive intonation) while others are anti-correlated (negative intonation). These two mechanisms can even be used in the same breath to produce complex changes in pitch. To our surprise, six of the ten USV types primarily used the positive intonation mechanism. These data support a novel and key role for the breathing system in the production of various types of vocalizations. In support of this hypothesis, we found re-activation of the breathing central pattern generator during the expiration phase of the vocal breath. This ectopic activity appeared as a ‘mini-breath’ (inspiratory diaphragm then post-inspiratory laryngeal muscle activities) nested within a normal expiration and corresponded with modulated airflow and pitch. This resembles how the vocalization central pattern generator, the iRO, patterns the rhythmic syllable structure of neonatal cry vocalizations. We show that the iRO is sufficient to induce most of the endogenous USV syllable types via the modulation of the breath airflow. In contrast to the natural lexicon, the pitch of the evoked USVs is primarily explained by positive intonation. These data imply that the iRO can produce the mechanism to pattern positive intonation, thereby suggesting that negative intonation derives from a separate neuronal component of the phonatory system. We propose that these two mechanisms can be used independently or in conjunction to generate the diverse repertoire of vocalizations (Fig. 5H).

The iRO likely patterns intonation for endogenous phonation

The description of the iRO within the adult neural circuit for phonation suggests a key role in patterning the endogenous adult vocalizations. In this case, we propose that the upstream periaqueductal gray input would “turn-on” the iRO which then co-opts the breathing pacemaker and coordinates its anti-phase activity with the laryngeal muscles to produce and pattern the changes in breath airflow and vocal pitch (positive intonation). The iRO can do this since it is presynaptic to both the breathing pacemaker and the laryngeal motor neurons. In this case, the brief re-activation of inspiratory muscles we observed would slow ongoing expiration, enabling a decrease in airflow speed, and thus pitch. After, relaxation of these muscles results in an increase in expiration airflow and pitch. This type of oscillatory modulation we observed has also been seen in neonatal cries generated by the iRO (Wei et al., 2022). An important next step will be to validate the necessity of the iRO in adult phonation, as anticipated from its necessary role in neonatal vocalization. None-the-less, the presence of the iRO across developmental stages implies a conserved role in innate vocalizations within the mouse and perhaps across the animal kingdom, where vocalization central pattern generators have been hypothesized and even identified in species from fish to birds to primates (Zhang et al., 2020, Chagnaud et al., 2011, Hage S.R., 2009b, Kelley et al., 2020).

The iRO can autonomously produce multiple vocalization patterns

A surprising finding is that ectopic activation of the iRO produces seven of the ten syllable types within the murine lexicon. How might this occur? One possibility is that the iRO has multiple modes which can each produce a different pattern of activity. Such a phenomenon has been demonstrated in other central pattern generating systems like the crustacean stomatogastric ganglia (Marder et al., 2001, Marder E., 2012). A more likely option is that additional mechanisms of vocal modulation are layered upon a basic pattern produced by the iRO. For example, other regions with direct control of the laryngeal motor neurons within RAm would add complexity to the vocalization induced by the iRO, akin to how vocal control by the human laryngeal motor cortex is conceptualized (Fig. 5H) (Dichter et al., 2018, Silva et al., 2022). Here we propose that perhaps just two mechanisms (breath airflow and laryngeal opening) account for the intricacy of the murine sounds produced, and the layering of these enables a basic pitch structure within a breath to become sophisticated. Of note, these resulting models do not simplistically explain the origin of the pitch jumps that are present in some USV types and these may instead arise from mechanisms like an active process to alter the conformation of the larynx or upper airway.

Recently, glutamatergic laryngeal premotor neurons in RAm were identified that are sufficient to elicit vocalizations, albeit somewhat abnormal, and also necessary to produce sound (Park et al., 2024, Veerakumar et al., 2023). This raises the possibility that these RAm neurons compose the cellular basis for the laryngeal control mechanism we propose produces negative intonation (Fig. 5H). However, the fact that ectopic excitation of these neurons or iRO elicits vocalizations suggests that reciprocal connections are engaged to create the dynamics of vocalizations. Consistently, iRO neurons project to RAm and vice versa (Fig. 3 and Park et al., 2024, Veerakumar et al., 2023). Interesting future studies will be to assess the necessity of either site to produce USVs upon ectopic excitation of the complementary region.

The control of breathing airflow is a novel biomechanical mechanism for intonation

Intonation is a key aspect of communication, whereby the same word or phrase could be used as a question or a statement simply by different fluctuations in pitch. Additionally, in tonal languages, the same sound with differences in pitch can have completely different meanings. Our findings describe a novel biophysical mechanism for intonation and a cellular basis. Now, the iRO or direct modulation of breathing can serve as a starting point to map higher level components of brain-wide vocalization circuits that structure additional subliminal layers of perception in speech.